---
title: "Implications of hyper-local Lassa virus transmission heterogeneity for vaccine trial site selection: a cross-sectional study in Nigeria, 2024"
format: 
  docx:
    reference-doc: reference-doc.docx
bibliography: references.bib
csl: eurosurveillance.csl
editor: source
---

```{r}
#| label: load-data
#| echo: false
#| include: false

if (!require("pacman")) install.packages("pacman")

pkgs <- c("brms",
          "cowplot",
          "flextable",
          "geodata",
          "ggridges",
          "ggsflabel",
          "ggspatial",
          "gt",
          "gtExtras",
          "gtsummary",
          "haven",
          "here",
          "janitor",
          "readxl",
          "sf",
          "smerc",
          "spdep",
          "terra",
          "tidybayes",
          "tidyverse",
          "tidyterra")

pacman::p_load(pkgs, character.only = T)

project_crs = "EPSG:4326"
utm_nigeria_CRS = "EPSG:26332"

colour_table <- read_rds(here("project_wide_data", "village_state.rds"))
state_colours <- setNames(unique(colour_table$state_colour), unique(colour_table$state))
village_colours <- setNames(colour_table$village_colour, colour_table$village)

scapes_spatial <- read_csv(here("project_wide_data", "scapes_spatial.csv")) %>%
  select(location, state, latitude, longitude) %>%
  vect(geom = c("longitude", "latitude"), crs = project_crs)

village_border_files <- list.files(here("project_wide_data", "village_borders"), pattern = "_border.shp$", full.names = TRUE)
village_borders <- lapply(village_border_files, function(x) {vect(x) |>
    project(y = utm_nigeria_CRS) |>
    aggregate()})
names(village_borders) <- c("dyegh", "enyandulogu", "ogamanna")
village_areas_km2 <- sapply(village_borders, function(v) {
  sum(terra::expanse(v) / 1000000)
})

hh_data <- read_rds(here("data", "hh_data.rds"))
i_data <- read_rds(here("data", "i_data.rds"))
serodata <- read_rds(here("data", "serology.rds"))

pop_estimates <- read_rds(here("project_wide_data", "village_pop_estimates.rds"))

village_state = tibble(abbreviation = c("zug", "dye", "iky", "oki", "oga", "ofo", "eze", "eny", "off"),
                       village = c("Zugu", "Dyegh", "Ikyogbakpev", "Okimbongha", "Ogamanna", "Ofonekom", "Ezeakataka", "Enyandulogu", "Offianka"),
                       lga = c("Vande Ikya", "Vande Ikya", "Vande Ikya", "Obubra", "Obubra", "Obubra", "Izzi", "Izzi", "Izzi"),
                       state = c("Benue [BN]", "Benue [BN]", "Benue [BN]", "Cross River [CR]", "Cross River [CR]", "Cross River [CR]", "Ebonyi [EB]", "Ebonyi [EB]", "Ebonyi [EB]"))

```

# Keywords
Lassa Fever; Lassa virus; Seroepidemiologic Studies; Zoonoses; Disease Reservoirs; Mastomys; Nigeria; Geography, Medical; Arenaviridae; Cross-Sectional Studies.

# Abstract

Introduction: Lassa virus (LASV) is a WHO priority pathogen with vaccine candidates advancing toward Phase III efficacy trials. Trial success depends on selecting sites with sufficient incidence, yet the local drivers of transmission remain poorly understood. We aimed to assess fine-scale variation in LASV exposure to inform trial site selection and control strategies.

Methods: We conducted a cross-sectional serosurvey in nine rural villages across three Nigerian states (Benue, Ebonyi, Cross River) during 2024. We tested 1,874 individuals for LASV IgG antibodies and analysed questionnaire data using Bayesian hierarchical models to identify demographic, environmental, and behavioural correlates of exposure. Spatial clustering was assessed using local Getis-Ord Gi* statistics.

Results: Overall seroprevalence was 3.3% but exhibited extreme heterogeneity, ranging from 1.7% to 5.1% between states and 0.8% to 6.5% between neighbouring villages. We found no consistent associations between seropositivity and self-reported risk factors, including rodent consumption or agricultural practices. Age-seroprevalence profiles differed markedly between locations, suggesting distinct local transmission histories (endemic vs. epidemic). Furthermore, we observed no significant village-wide spatial clustering of seropositive households.

Discussion: LASV exposure is driven by complex, hyper-local factors that are not captured by broad regional risk mapping. The absence of clear demographic risk profiles challenges the efficacy of spatially targeted interventions based on passive surveillance. For vaccine trial planning, relying on regional notification rates may lead to the selection of sites with insufficient transmission intensity. We recommend granular, active One Health surveillance—integrating human serology with longitudinal reservoir monitoring—to verify local transmission dynamics prior to trial enrolment.

# Key Public Health Messages

## What did you want to address in this study and why?  
Lassa fever is a dangerous viral disease in West Africa. Scientists are currently developing vaccines, but to prove these vaccines work, they must conduct clinical trials in villages where the virus is actively spreading and cases accrue. We wanted to check if current risk maps—which predict high-risk areas based on broad geography—are accurate enough to help researchers choose the right villages for these important trials.

## What have we learnt from this study? 
We tested over 1,800 people in rural Nigeria and found that Lassa fever infection rates vary substantially between neighbouring villages. One village might have many past infections (up to 6.5%), while a village just a few kilometres away has almost none (0.8%). Surprisingly, we found that common risk factors, such as farming or eating rodents, did not predict which villages had high infection rates.

What are the implications of your findings for public health? 
Current broad-scale maps are not detailed enough to predict local risk. If vaccine researchers rely on these maps, they might accidentally choose trial sites with very little virus transmission, leading to trials not reaching clinical endpoints and failing. We recommend that before starting expensive clinical trials, researchers must actively monitor both local people and rodent populations to confirm the virus is circulating in potential trial locations.

# Introduction

Lassa virus (*Mammarenavirus lassaense*; Arenaviridae, LASV) is a zoonotic arenavirus endemic to West Africa, with regular outbreaks occurring in Nigeria, Guinea, Liberia, and Sierra Leone. It causes Lassa fever, a viral haemorrhagic illness responsible for an estimated 900,000 annual infections and over 200 deaths annually in Nigeria alone [@basinskiBridgingGapUsing2021]. Clinical presentation ranges from asymptomatic infection to severe haemorrhagic fever with case fatality exceeding 20% in hospitalized patients, for whom treatment options remain limited [@salamRibavirinTreatingLassa2022].

The primary reservoir for LASV is *Mastomys natalensis*, a highly fecund, synanthropic rodent widely distributed across sub-Saharan Africa [@bellocqDhatiWelelVirus2020]. It thrives in agricultural and peri-domestic environments, creating substantial opportunities for human contact [@marienHouseholdsHotspotsLassa2020]. Infected rodents, particularly those infected vertically in utero, can become persistent shedders of the virus in excreta, while its fluctuating population dynamics influence seasonal transmission intensity [@hoffmannLassaVirusPersistence2024; @leirsTwentynineYearsContinuous2023].

Human infection occurs primarily through direct or indirect contact with infected rodents, their excreta, contaminated food, or aerosolised particles [@bauschLassaFeverGuinea2001]. Occupational and domestic exposures, particularly in farming, food storage, and rodent hunting contexts, are commonly implicated. Critically, seroprevalence surveys show that exposure is widespread, with most infections (~80%) being asymptomatic or subclinical; consequently, passive surveillance-derived case counts severely underestimate the true infection burden [@simonsLassaFeverCases2022c].

Environmental and ecological conditions strongly influence LASV spillover [@reddingEnvironmentalmechanisticModellingImpact2016]. Land use changes like deforestation and agricultural intensification can increase human-rodent contact, while shifts in rodent community composition may also moderate risk [@banguraLassaVirusCirculation2021; @eskewReservoirDisplacementInvasive2024; @simonsLandUseGradients2025]. Despite these broad-scale drivers, observed variation in LASV prevalence between ecologically similar areas highlights the critical importance of fine-scale factors [@Grant2023-em]. These include social conditions, such as household structure, food storage methods, and specific human-animal interactions.

Reliable epidemiological data are urgently needed to support the development of LF therapeutics and vaccines. LASV is a designated priority pathogen for the World Health Organization and the Coalition for Epidemic Preparedness Innovations (CEPI), with multiple vaccine candidates currently advancing into Phase II clinical trials [@malkinSafetyImmunogenicityRVSV2025]. The success of future Phase III efficacy trials relies heavily on identifying hotspot communities where incidence is sufficient to accrue clinical cases to measure a vaccine effect. However, relying on broad-scale incidence data to select these sites risks recruitment failure if transmission is more focal than regional risk maps suggest.

Current Lassa fever surveillance relies primarily on passive case detection at sentinel hospitals, which overlooks mild infections and asymptomatic seroconversion in the community [@marienSerosurveillanceIdentifiesEndemic2025a]. As a result, major gaps remain in understanding why transmission risk varies so markedly at local scales, where the conditions facilitating spillover remain poorly defined [@friantScalingLassaVirus2025]. To address this, we conducted a cross-sectional study in nine communities across three Nigerian states. We combined serological testing with structured questionnaires to investigate fine-scale drivers of exposure. Our objectives were to: 1) estimate LASV seroprevalence at state and village levels; 2) characterise individual and household correlates of seropositivity; and 3) explore spatial heterogeneity in local transmission risk.

# Materials and Methods

The primary objective of this study was to estimate the seroprevalence of LASV IgG in nine communities across three Nigerian states. Secondary objectives were to explore demographic, environmental, behavioural, and occupational correlates of seropositivity and to assess fine-scale spatial heterogeneity of exposure.

## Study design and participants

This cross-sectional baseline survey was conducted as part of the SCAPES longitudinal study [@friantScalingLassaVirus2025]. Nine villages across three states (Benue [BN], Ebonyi [EB], and Cross River [CR]) were selected using a tool incorporating population estimates, land cover (as a proxy for *M. natalensis* habitat suitability), and logistical accessibility [@imirzianSitetoolApplicationField2025]. Final selection was confirmed through site visits and consultation with local public health surveillance teams.

Households were enumerated to estimate population size and plan recruitment. Villages ranged from 110 to 321 households (population 465–2,313; Supplementary Materials Section 1). We aimed to recruit ~20% of households per village (target ~70), enrolling four individuals per household: an adult male, adult female, adolescent (12–18), and a randomly selected child (<12). Children under 12 provided dried blood spot (DBS) samples but did not complete questionnaires.

Based on published LASV seroprevalence data [@Kerneis2009-sb; @Grant2023-em; @marienSerosurveillanceIdentifiesEndemic2025a], we assumed high-risk (>40% seroprevalence), medium-risk (10–20%), and low-risk (<5%) village categories. We estimated that 180 individuals would provide 80% power (α = 0.05, β = 0.2) to detect significant differences in LASV seroprevalence between village categories. Household-level clustering of serostatus was expected; hence, the design was powered to detect differences at both village and household levels.

Systematic sampling was used to select households within each village. Routes were designed to approach every *n*th household (where *n* is the total number of households divided by the target sample size). Three households (all in BN) declined participation; in these instances, the next household was enrolled. The participant flow diagram is available in Supplementary Materials Section 2.

## Ethics approval

The study protocol was approved by the Institutional Review Board of the Pennsylvania State University, USA (STUDY00019989), and by the Nigerian National Health Research Ethics Committee (AEC/03/168/24). Additional approval was obtained from the health research ethics committees of the public health agencies in each participating state. The study was conducted in accordance with the Declaration of Helsinki. Enrolment involved engagement with community leaders. Informed consent was obtained from all adults; assent and parental consent were obtained for minors.

## Data collection

A questionnaire covering demographic, environmental, behavioural, and occupational domains was administered [@RiskLabPSUBaselineseroprevpublic2025]. DBS samples were collected by fingerstick on Whatman 903 cards, air-dried, and transported with desiccant. Samples were eluted and tested for LASV IgG using the Panadea LASV ELISA kit following the manufacturer's protocol modified for DBS [@Longet2023-hk; @Soubrier2022-gr].

## Statistical methods

Analyses were performed in R (v4.2.3) [@R]. Questionnaire responses were linked to individual and household records using unique identifiers. Household GPS coordinates were used for spatial analyses. Data cleaning involved range validation, logical consistency checks, and outlier flagging. Missingness was negligible (<1%); analyses were restricted to complete cases.
### Descriptive group comparisons

To compare characteristics between villages or states, we fitted separate Bayesian models using the `brms` package [@brms] (Gaussian linear models for continuous variables; multinomial logistic regression for categorical). We defined a notable difference between groups as evidence that the variability of the outcome between villages/states was non-zero. This was assessed by examining the posterior distribution of the random effect standard deviation (SD) parameter for the grouping variable. A notable difference was concluded if the lower bound of the 95% Credible Interval (CrI) for the SD parameter was >0.1, indicating non-negligible variation.

### Estimating LASV seroprevalence

We estimated seroprevalence and 95% CrIs using Bayesian generalised linear mixed models (GLMM) with a binary outcome (seropositive/seronegative) and a single fixed effect for state or village. Posterior predictions generated marginal estimates averaged over sampled individuals. To explore age-related variation, we fitted a Bayesian generalised additive model (GAM) with village-specific smooth terms for age. Posterior expected probabilities were calculated across a grid of age values to obtain modelled estimates. The full posterior summaries (central estimates and CrIs) for these model comparisons are presented in Supplementary Materials. Differences classed as important are indicated in the main text and tables.

### Characterising behavioural and exposure correlates of seropositivity

We investigated associations between LASV seropositivity and a range of *a priori* risk factors using a Bayesian GLMM framework, specifying a Bernoulli (logit) likelihood for serostatus as the binary outcome. Analyses were conducted at both individual and household levels, with explanatory variables grouped into four conceptual domains: demographic, environmental, behavioural, and occupational.

To account for potential non-independence of observations within villages, reflecting shared environmental, infrastructural, or cultural exposures, a random intercept for village was included in all models. Univariable models were fitted separately for each covariate to examine associations in isolation, adjusting only for village-level clustering. We evaluated the magnitude and uncertainty of potential risk factors by examining the posterior odds ratios (OR) and their 95% CrIs. An interval that included 1.0 was interpreted as showing no strong evidence for an association.

### Exploring spatial heterogeneity in LASV exposure

To assess spatial patterns in LASV exposure at the household level, we geocoded each sampled household and identified those with at least one seropositive individual. We first applied Global Moran’s I to test for overall spatial autocorrelation of household seropositivity within each village. To identify potential micro-scale clustering, we then used local Getis-Ord Gi* analysis using the `spdep` package, which compares the value at each household (the proportion of seropositive members) to the values of neighbouring households within a defined spatial lag (300m), generating a z-score indicating statistically significant hot spots (high-value clusters) and cold spots (low-value clusters) [@spdep].

As a *post hoc* check on the relationship between hotspot classification and seropositivity, we fit a simplified logistic regression model. This model used household serostatus (presence/absence of seropositive individuals) as the outcome and compared two groups: households located in any Gi∗-defined cluster (Hotspot or Coldspot) versus those classified as Not Significant. This allowed us to robustly quantify the association between being in a statistically-defined cluster and the household’s observed seropositivity.

Given the low seroprevalence and relatively small village area (median = 6.7km^2^), these analyses were intended as descriptive and exploratory, with a focus on identifying potential micro-scale heterogeneity rather than formally testing for clustering.

# Results

## Study population

Between 16 December 2023 and 22 July 2024, we enrolled 1,926 individuals (1,874 obtained DBS samples, 1,469 completed questionnaires) from 577 households across nine villages in three Nigerian states (BN, CR, EB). This sample represented 27% of households and 11% of the expected total village population (Table 1, Table 2 and Supplementary Table 1). 

## Households

Household composition and infrastructure showed notable variation across the study villages, particularly in household structure (Table 1). There was strong evidence of differences in household size, ranging from median sizes of 8 in Dyegh [BN] to 4 in Ofonekom [CR]. Similarly, the number of buildings per household and number of single-room buildings differed substantially between villages, with households in BN generally being larger and comprising more buildings. 

Environmental characteristics were more consistent across sites. Proximity to bush (51%) and farms (66%), predominant sanitation method (field defecation: 80%), and rodent entry into homes (94%) did not show strong evidence of differing between villages.

Households employed various methods for rodent removal (Supplementary Materials Table 1). In contrast to the uniformity of rodent entry, our analysis found strong evidence that the specific methods used, such as poison (73%), sticks (44%), and traps, varied substantially between villages. However, the ultimate outcome for captured rodents was largely consistent; 88% of households reported disposing of them. While disposal was uniform, the practice of consuming captured rodents did show notable variation between villages, consistent with the individual-level data (Table 2).

```{r}
#| label: prep-data-table-1
#| echo: false
#| cache: true
#| warning: false
#| include: false

tbl_1_data <- hh_data |>
  select(`_index`, household_id, n_people, compound, n_single_room, n_multi_room, rodent_removal_home_method, rodent_mitigation_method, rodent_remove_use, livestock, toilet, storage_cooked_containers, storage_packaged_containers, storage_seed_crop_containers_c, storage_seed_crop_containers_g, latitude, longitude, surroundings_household, surroundings_outside_other) |>
  mutate(rodents_enter = case_when(is.na(rodent_removal_home_method) ~ "No",
                                   TRUE ~ "Yes"),
         proximity_bush = case_when(str_detect(surroundings_household, "bush") ~ "Yes",
                                    TRUE ~ "No"),
         proximity_farm = case_when(str_detect(surroundings_household, "farm") ~ "Yes",
                                    TRUE ~ "No"),
         toilet_use = fct(case_when(str_detect(toilet, "toilet_plumbing") ~ "Plumbed toilet",
                                    str_detect(toilet, "pit_latrine") ~ "Pit latrine",
                                    str_detect(toilet, "open_system") ~ "Open system",
                                    str_detect(toilet, "field_defecation") ~ "Field defecation",
                                    TRUE ~ "Other"),
                          levels = c("Plumbed toilet", "Pit latrine", "Open system", "Field defecation", "Other")),
         rodent_control_animals = case_when(str_detect(rodent_mitigation_method, "cats|dogs") ~ "Yes",
                                            TRUE ~ "No"),
         rodent_control_containers = case_when(str_detect(rodent_mitigation_method, "containers") ~ "Yes",
                                               TRUE ~ "No"),
         rodent_control_burrows = case_when(str_detect(rodent_mitigation_method, "seal_burrows") ~ "Yes",
                                            TRUE ~ "No"),
         rodent_control_structure = case_when(str_detect(rodent_mitigation_method, "structure|convert") ~ "Yes",
                                              TRUE ~ "No"),
         rodent_remove_dispose = case_when(str_detect(rodent_remove_use, "dispose_them") ~ "Yes",
                                           TRUE ~ "No"),
         rodent_remove_eat = case_when(str_detect(rodent_remove_use, "eat_them") ~ "Yes",
                                       TRUE ~ "No"),
         rodent_remove_feed = case_when(str_detect(rodent_remove_use, "feed_them_to_animals") ~ "Yes",
                                        TRUE ~ "No"),
         rodent_remove_sell = case_when(str_detect(rodent_remove_use, "sell_them") ~ "Yes",
                                        TRUE ~ "No"),
         rodent_method_animal = case_when(str_detect(rodent_removal_home_method, "cat|dog") ~ "Yes",
                                          TRUE ~ "No"),
         rodent_method_trap = case_when(str_detect(rodent_removal_home_method, "gumtrap|traps") ~ "Yes",
                                        TRUE ~ "No"),
         rodent_method_poison = case_when(str_detect(rodent_removal_home_method, "poison") ~ "Yes",
                                          TRUE ~ "No"),
         rodent_method_sticks = case_when(str_detect(rodent_removal_home_method, "sticks") ~ "Yes",
                                          TRUE ~ "No")
  )  |>
  rowwise() |>
  mutate(n_buildings = sum(n_single_room, n_multi_room, na.rm = TRUE)) |>
  select(`_index`, household_id, n_people, n_buildings, n_single_room, n_multi_room, proximity_bush, proximity_farm, toilet_use, rodents_enter, rodent_control_animals, rodent_control_burrows, rodent_control_structure, rodent_control_containers,
         rodent_method_animal, rodent_method_poison, rodent_method_sticks, rodent_method_trap, rodent_remove_dispose, rodent_remove_eat, rodent_remove_feed, rodent_remove_sell) %>%
  left_join(serodata |>
              mutate(household_id = str_remove(as.character(id), "-\\d+$"),
                     is_positive = interpretation == "Positive") |>
              group_by(household_id) |>
              summarise(n_positive = sum(is_positive, na.rm = TRUE)),
            by = "household_id") |>
  mutate(abbreviation = str_split(household_id, "-", simplify = TRUE)[, 1]) |>
  left_join(village_state, by = "abbreviation") |>
  mutate(village = fct(village, levels = c(village_state$village)),
         n_single_room = as.integer(n_single_room),
         n_multi_room = as.integer(n_multi_room))

bayesian_results <- list(
  n_people = TRUE,
  n_buildings = TRUE,
  n_single_room = TRUE,
  proximity_bush = TRUE,
  proximity_farm = TRUE,
  toilet_use = TRUE,
  rodents_enter = FALSE,
  rodent_method_animal = TRUE,
  rodent_method_poison = TRUE,
  rodent_method_sticks = TRUE,
  rodent_method_trap = TRUE,
  rodent_remove_dispose = FALSE,
  rodent_remove_eat = TRUE,
  rodent_remove_feed = FALSE,
  rodent_remove_sell = FALSE
)
```


```{r}
#| label: create-table-1
#| echo: false
#| tbl-cap: "Household-level demographic and environmental characteristics"

tbl_summary_input_main <- tbl_1_data |>
  mutate(any_seropositive = factor(case_when(n_positive == 0 ~ "All negative", n_positive >= 1 ~ "At least one positive"), levels = c("All negative", "At least one positive"))) |>
  select(
    village,
    n_people,
    any_seropositive,
    n_buildings,
    n_single_room,
    proximity_bush,
    proximity_farm,
    toilet_use,
    rodents_enter
  )

household_summary_main <- tbl_summary(
  data = tbl_summary_input_main,
  by = village,
  statistic = list(
    all_continuous() ~ "{median} ({IQR})",
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = all_continuous() ~ 1,
  # Group variables for readability
  label = list(
    n_people ~ "Household size (individuals)",
    any_seropositive ~ "Household seropositivity",
    n_buildings ~ "Number of buildings",
    n_single_room ~ "Single-room buildings",
    proximity_bush ~ "Proximity to bush",
    proximity_farm ~ "Proximity to farm",
    toilet_use ~ "Type of toilet",
    rodents_enter ~ "Rodents enter home"
  ),
  missing = "no"
) |>
  add_overall() |>
  add_n()

tbl_with_symbols <- household_summary_main
for (var in names(bayesian_results)) {
  if (isTRUE(bayesian_results[[var]])) {
    tbl_with_symbols <- tbl_with_symbols |>
      # Modify the label for the specific variable row
      modify_table_body(
        ~ .x %>%
          dplyr::mutate(
            label = ifelse(variable == var, paste0(label, "*"), label)
          )
      )
  }
}

hh_row <- tibble(
  variable = "n_hh",
  var_type = "character",
  row_type = "label",
  var_label = "Number of households",
  label = "Number of households",
  n = NA_character_,
  stat_0 = NA_character_,
  stat_1 = "205 (31%)",   # Benue
  stat_2 = "305 (21%)",
  stat_3 = "149 (43%)",
  stat_4 = "191 (34%)",   # Cross River
  stat_5 = "231 (29%)",
  stat_6 = "110 (55%)",
  stat_7 = "321 (21%)",   # Ebonyi
  stat_8 = "281 (23%)",
  stat_9 = "381 (17%)"
)

tbl_with_symbols$table_body <- bind_rows(hh_row, tbl_with_symbols$table_body)

tbl_1_final_revised <- tbl_with_symbols |>
  modify_header(label ~ "**Characteristic**") |>
  bold_labels() |>
  # Add the spanning headers to group villages by state
  modify_spanning_header(c("stat_1", "stat_2", "stat_3") ~ "**Benue [BN]**",
                         c("stat_4", "stat_5", "stat_6") ~ "**Cross River [CR]**",
                         c("stat_7", "stat_8", "stat_9") ~ "**Ebonyi [EB]**") |>
  # Add the Bayesian model footnote (Applies to the 'label' column)
  modify_footnote_body(
    footnote = "*Indicates strong evidence of a difference between villages, as determined by a Bayesian model. See supplementary for central estimates and 95% CrI.",
    columns = label,
    rows = str_detect(label, "\\*") 
  ) |>
  modify_footnote_body(
    footnote = "% refers to proportion of households enrolled from all households in the village",
    columns = label,
    rows = label == "Number of households"
  ) |>
  modify_footnote_body(
    footnote = "Maximum of 4 household members were tested irrespective of household size",
    columns = label,
    rows = label == "Household seropositivity"
  ) |>
  as_gt() |>
    gt::tab_style(
    style = gt::cell_fill(color = "lightgray"),
    locations = gt::cells_body(
      columns = c("stat_1", "stat_2", "stat_3", "stat_7", "stat_8", "stat_9")
    )
  ) |>
  tab_options(
    table.border.top.style = "solid",
    table.border.bottom.style = "solid",
    column_labels.border.bottom.style = "solid",
    table_body.vlines.style = "none",
    heading.border.bottom.style = "none"
  )

tbl_1_final_revised
```

## Individuals

Demographics and behaviours varied notably across sites (Table 2). Migration patterns differed (Strong Evidence); the percentage of residents born in their current village was substantially lower in BN compared to CR. However, most participants were permanent residents (median 12 months/year).  A balanced distribution by sex was achieved by the sampling design for all sites. Median age (Overall: 28, IQR: 31) was consistent across villages.

Educational attainment varied substantialy; participants in BN and CR had higher rates of secondary education completion compared to EB. Subsistence farming was the most common occupation (59%), followed by trading (17%) and hired agricultural work (6%). Primary occupations did not show strong evidence of difference between villages.

Rodent consumption practices differed substantially. Current rodent consumption was common in most BN and CR villages (e.g., 91% in Dyegh [BN] and 73% in Ogamanna [CR]) but was rarer in the EB villages (e.g., 4.5% in Enyandulogu). Even among those not currently consuming rodents, the proportion who reported past consumption also varied, being generally higher in EB villages (e.g., 78% in Offianka) than elsewhere. Indirect rodent contact, such as cleaning rodent excreta, was nearly universal (99%) and did not differ between villages.

```{r}
#| label: prep-table-2-data
#| cache: true
#| echo: false
#| include: false

tbl_2_data <- i_data |>
  select(id, age, sex, ethnicity, religion, education, village_residence, months_residence_in_year, community_pob, income_individual, income, field_entry, field_work_freq, forest_entry, forest_entry_freq, current_rodent_consumption, past_rodent_consumption,
         rodent_eat_reason, excreta_cleaning, rodent_diseases) |>
  left_join(serodata,
            by = "id")
tbl_2_data$sex[tbl_2_data$id == "dye-6-1"] <- "Male"

income_dummies <- tbl_2_data |>
  select(id, income) |>
  separate_rows(income, sep = " ") |>
  mutate(income = str_trim(income)) |>
  dplyr::distinct(id, income) |>
  mutate(present = 1) |>
  pivot_wider(names_from = income, values_from = present, values_fill = 0) |>
  mutate(ag_worker = case_when(ag_worker == 0 & ag_worker_elsewhere == 1 ~ 1,
                               TRUE ~ ag_worker))|>
  select(-`NA`, -ag_worker_elsewhere)

rodent_eat_reason_dummies <- tbl_2_data |>
  select(id, rodent_eat_reason) |>
  separate_rows(rodent_eat_reason, sep = " ") |>
  mutate(rodent_eat_reason = str_trim(rodent_eat_reason)) |>
  dplyr::distinct(id, rodent_eat_reason) |>
  mutate(present = 1) |>
  pivot_wider(names_from = rodent_eat_reason, values_from = present, values_fill = 0) |>
  select(-`NA`)

tbl_2_data_processed <- tbl_2_data %>%
  left_join(income_dummies, by = "id") %>%
  left_join(rodent_eat_reason_dummies, by = "id") %>%
  mutate(village_residence = factor(village_residence, levels = c(village_state$village)),
         past_rodent_consumption = case_when(current_rodent_consumption == "Yes" ~ NA,
                                             TRUE ~ past_rodent_consumption),
         education = factor(education, levels = c("None/No formal schooling",
                                                  "Primary education (up until age 12)",
                                                  "Junior secondary education (up until age 16)",
                                                  "Senior secondary education (up until age 19)",
                                                  "University level education",
                                                  "Master's, Doctoral level degree",
                                                  "Other")))

bayesian_results_individual <- list(
  # Demographics & Residency
  age = FALSE,
  sex = FALSE,
  education = TRUE,
  months_residence_in_year = FALSE,
  community_pob = TRUE,
  
  # Occupations
  farming = FALSE,
  ag_worker = FALSE,
  
  # Behaviours
  field_entry = FALSE,
  forest_entry = TRUE,
  current_rodent_consumption = TRUE,
  past_rodent_consumption = TRUE,
  excreta_cleaning = FALSE,
  
  # Reasons for Rodent Consumption
  taste = FALSE,
  availability = FALSE,
  cheap = FALSE,
  nutrition = FALSE,
  cultural = FALSE
)

```

```{r}
#| label: create-table-2
#| echo: false
#| warning: false
#| tbl-cap: "Individual-level demographic, occupational, and behavioral characteristics"

tbl_2_summary_input_condensed <- tbl_2_data_processed |>
  drop_na(interpretation, age, sex) |>
  mutate(
    # Merge Equivocal into Negative
    interpretation = case_when(
      interpretation == "Equivocal" ~ "Negative",
      TRUE ~ interpretation
    ),
    # Simplify Education
    education_four_levels = case_when(
      str_detect(education, "None") ~ "None",
      str_detect(education, "Primary") ~ "Primary",
      str_detect(education, "Junior|Senior") ~ "Secondary",
      str_detect(education, "University|Master|Other") ~ "Post-secondary",
      is.na(education) ~ NA_character_
    ),
    education_four_levels = factor(education_four_levels, levels = c("None", "Primary", "Secondary", "Post-secondary")),
    # Handle Excreta Cleaning logic
    excreta_cleaning = case_when(
      is.na(excreta_cleaning) & age >= 12 ~ "No",
      is.na(excreta_cleaning) ~ NA_character_,
      TRUE ~ "Yes"
    )
  ) |>
  select(
    village_residence,
    interpretation, 
    age, 
    sex,
    education_four_levels,
    community_pob,
    farming, ag_worker, trader,
    current_rodent_consumption,
    past_rodent_consumption,
    excreta_cleaning  
  )

individual_summary_condensed <- tbl_summary(
  data = tbl_2_summary_input_condensed,
  by = village_residence,
  statistic = list(
    all_continuous() ~ "{median} ({IQR})",
    all_categorical() ~ "{n} ({p}%)"
  ),
  digits = list(all_continuous() ~ 0),
  missing = "no",
  # Value labels to filter specific rows (e.g., only show Female)
  value = list(
    sex ~ "Female",
    community_pob ~ "Yes",
    farming ~ "1",
    ag_worker ~ "1",
    trader ~ "1",
    current_rodent_consumption ~ "Yes",
    past_rodent_consumption ~ "Yes",
    excreta_cleaning ~ "Yes",
    interpretation ~ "Positive"
  ),
  label = list(
    interpretation ~ "LASV Seropositive",
    age ~ "Age (years)",
    sex ~ "Sex: Female",
    education_four_levels ~ "Education*",
    community_pob ~ "Born in study village*",
    farming ~ "Occupation: Farmer",
    ag_worker ~ "Occupation: Agricultural worker",
    trader ~ "Occupation: Trader",
    current_rodent_consumption ~ "Rodent consumption*",
    past_rodent_consumption ~ "Past rodent consumption only*",
    excreta_cleaning ~ "Cleaned rodent excreta"
  )
) |>
  add_overall() |>
  add_n()

pop_row <- tibble(
  variable = "village_population",
  test_name = NA_character_,
  var_type = "character",
  row_type = "label",
  var_label = "Estimated village population",
  label = "Estimated village population",
  n = NA_character_,
  stat_0 = NA_character_,
  stat_1 = "1,227 (15%)", stat_2 = "930 (23%)", stat_3 = "2,242 (9%)", # BN
  stat_4 = "1,196 (21%)", stat_5 = "1,401 (16%)", stat_6 = "464 (40%)",  # CR
  stat_7 = "2,125 (11%)", stat_8 = "1,790 (12%)", stat_9 = "2,310 (9%)"  # EB
)

individual_summary_condensed$table_body <- bind_rows(pop_row, individual_summary_condensed$table_body)

# 4. Final Formatting (Eurosurveillance Style)
tbl_2_final_condensed <- individual_summary_condensed |>
  modify_header(label ~ "**Characteristic**") |>
  bold_labels() |>
  modify_spanning_header(
    c("stat_1", "stat_2", "stat_3") ~ "**Benue [BN]**",
    c("stat_4", "stat_5", "stat_6") ~ "**Cross River [CR]**",
    c("stat_7", "stat_8", "stat_9") ~ "**Ebonyi [EB]**"
  ) |>
  modify_footnote_body(
    footnote = "*Indicates strong evidence of a difference between villages (Bayesian hierarchical model).",
    columns = label,
    rows = str_detect(label, "\\*") 
  ) |>
  as_gt() |>
  gt::tab_style(
    style = gt::cell_fill(color = "lightgrey"),
    locations = gt::cells_body(columns = c("stat_1", "stat_2", "stat_3", "stat_7", "stat_8", "stat_9"))
  ) |>
  tab_options(table_body.vlines.style = "none")

tbl_2_final_condensed
```


## LASV seroprevalence estimates

Of the 1,874 individuals tested, 61 were seropositive for LASV IgG, yielding a crude overall seroprevalence of 3.3%. These individuals belonged to 59 households, with only two households (both in Ikyogbakpev [BN]) containing more than one seropositive member. No seropositive individuals reported a prior diagnosis of Lassa fever.

Model-based estimates revealed marked heterogeneity in exposure risk across the study area (Table 3). The overall seroprevalence was 3.17% (95% CrI: 2.47-4.04%). This varied at the state level (CR = 5.13% [3.66–6.97%]; BN = 2.63% [1.56–4.06%]; EB = 1.65% [0.86–2.85%]) and showed even greater variation at the village level (ranging from Ezeakataka [EB] = 0.79% [0.13-2.45%] to Okimbongha [CR] = 6.48% [3.9-9.97%]) (Table 3.).

```{r}
#| label: loading-estimated-prevalence
#| echo: false
#| warning: false
#| message: false
output_dir <- here("output", "models")
village_prevalence_draws <- readRDS(here(output_dir, "village_prevalence_draws.rds"))
state_prevalence_draws <- readRDS(here(output_dir, "state_prevalence_draws.rds"))
age_seroprevalence_data <- readRDS(here(output_dir, "age_seroprevalence_data.rds"))
```

```{r}
#| label: table-3
#| echo: false
#| warning: false
#| tbl-cap: "Model-based LASV IgG seroprevalence estimates at the state and village level."

village_agg_ns <- serodata |>
  mutate(abbreviation = str_split(id, pattern = "-", simplify = TRUE)[, 1],
         y = case_when(interpretation == "Positive" ~ 1, TRUE ~ 0)) |>
  left_join(village_state, by = "abbreviation") |>
  group_by(village) |>
  summarise(
    `N sampled` = n(), 
    `N positive` = sum(y, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(Group = village)

state_agg_ns <- village_agg_ns |>
  mutate(state = case_when(
    Group %in% c("Dyegh", "Ikyogbakpev", "Zugu") ~ "Benue",
    Group %in% c("Ofonekom", "Ogamanna", "Okimbongha") ~ "Cross River",
    Group %in% c("Enyandulogu", "Ezeakataka", "Offianka") ~ "Ebonyi"
  )) |>
  group_by(state) |>
  summarise(
    `N sampled` = sum(`N sampled`), 
    `N positive` = sum(`N positive`),
    .groups = "drop"
  ) |>
  rename(Group = state)

all_ns <- bind_rows(village_agg_ns, state_agg_ns)

combined_draws <- bind_rows(
  village_prevalence_draws |> rename(Group = village) |> mutate(level = "village"),
  state_prevalence_draws   |> rename(Group = state)   |> mutate(level = "state")
) |>
  mutate(Group = case_when(
    str_detect(Group, "CrossRiver") ~ "Cross River",
    TRUE ~ Group
  ))

table_3_input <- combined_draws |>
  group_by(Group, level) |>
  summarise(
    median_prevalence = median(prevalence * 100),
    lower_ci = quantile(prevalence * 100, 0.025),
    upper_ci = quantile(prevalence * 100, 0.975),
    .groups = "drop"
  ) |>
  left_join(all_ns, by = "Group") |>
  mutate(
    `Pos` = `N positive`,
    `Prev` = median_prevalence,
    `Low` = lower_ci,
    `High` = upper_ci
  )

display_order <- c(
  "Benue", "Dyegh", "Ikyogbakpev", "Zugu",
  "Cross River", "Ofonekom", "Ogamanna", "Okimbongha",
  "Ebonyi", "Enyandulogu", "Ezeakataka", "Offianka"
)

gt_table_revised <- table_3_input |>
  mutate(Group = factor(Group, levels = display_order)) |>
  arrange(Group) |>
  select(Group, level, `N sampled`, `Pos`, `Prev`, `Low`, `High`) |>
  gt() |>
  cols_label(
    Group = "State/Village",
    `N sampled` = "N",
    `Pos` = "n",
    `Prev` = "%",
    `Low` = "Lower",
    `High` = "Upper"
  ) |>
  tab_spanner(label = "Seropositive", columns = c(`Pos`, `Prev`)) |>
  tab_spanner(label = "95% CrI", columns = c(`Low`, `High`)) |>
  fmt_number(columns = c(`Prev`, `Low`, `High`), decimals = 2) |>
  cols_align(align = "left", columns = Group) |>
  # Bold State Rows
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Group, rows = level == "state")
  ) |>
  # Indent Village Rows
  tab_style(
    style = cell_text(indent = px(20)),
    locations = cells_body(columns = Group, rows = level == "village")
  ) |>
  cols_hide(columns = level) |>
  tab_options(table_body.vlines.style = "none")

gt_table_revised
```


The relationship between age and seropositivity was not uniform, revealing distinct, village-specific transmission dynamics (Figure 1). Two villages (Ogamanna [CR] and Ofonekom [CR]) exhibited the classic pattern of increasing seroprevalence with age, consistent with long-term, cumulative exposure. This trend was statistically robust in Ogamanna [CR] (Posterior probability of increase > 99%), while Ofonekom [CR] showed a similar but more uncertain trajectory (Probability ~76%). Notably, these two villages are geographically adjacent (< 5km apart), suggesting a shared local ecology. Conversely, Okimbongha [CR] and Ezeakataka [EB] displayed profiles peaking in younger groups before declining, suggestive of recent, focal exposure, though statistical evidence for the decline was weaker (Probability 60–66%). In remaining villages, seroprevalence remained consistently low.

```{r}
#| label: produce-age-seroprev
#| echo: false
#| warning: false

age_seroprev <- age_seroprevalence_data |>
  group_by(age, village) |>
  median_qi(.epred * 100) |>
  left_join(village_state, by = "village") |>
  mutate(village = factor(village, levels = c(village_state$village)),
         state = factor(state, levels = c(unique(village_state$state)))) |>
  ggplot(aes(x = age, y = `.epred * 100`, colour = village)) +
  geom_line(linewidth = 1.2) +
  # Use the .lower and .upper columns for the ribbon
  geom_ribbon(aes(ymin = .lower, ymax = .upper, fill = village), alpha = 0.2, color = NA) +
  scale_colour_manual(values = village_colours) +
  scale_fill_manual(values = village_colours) +
  scale_x_continuous(limits = c(0, 70)) +
  scale_y_continuous(limits = c(0,35)) +
  labs(
    colour = "Village",
    fill = "Village",
    x = "Age (years)",
    y = "Estimated Seroprevalence (%)"
  ) +
  theme_minimal() +
  facet_grid(~ state, scales = "fixed")

save_plot(plot = age_seroprev, filename = here("output", "figure_1.pdf"), base_width = 8, bg = "white")

```

```{r}
#| label: compare-age-seroprev
#| echo: false
#| warning: false

age_young <- 10
age_old <- 60

trend_check <- age_seroprevalence_data |>
  filter(age %in% c(age_young, age_old)) |>
  ungroup() |>
  select(village, .draw, age, .epred, -.row) |>
  pivot_wider(names_from = age, values_from = .epred, names_prefix = "age_") |>
  # Calculate the difference for every single posterior draw
  mutate(
    diff_increasing = age_60 - age_10, # Positive if increasing
    diff_declining  = age_10 - age_60  # Positive if declining
  ) |>
  group_by(village) |>
  summarise(
    # What % of draws show an increase?
    prob_increasing = mean(diff_increasing > 0),
    # What % of draws show a decline?
    prob_declining = mean(diff_declining > 0),
    # Get the median difference
    median_diff = median(diff_increasing)
  )

```


```{r}
#| label: plot-age-seroprev
#| fig.cap: "Village-specific age-seroprevalence curves for Lassa virus. Lines represent median posterior estimates from a generalized additive model, and ribbons show the 95% credible interval."
#| echo: false

knitr::include_graphics(here("output", "figure_1.pdf"))

```

## Risk factors for seropositivity

```{r}
#| label: univariable-analyses
#| eval: true
#| message: false
#| warning: false
#| include: false

i_data_models <- tbl_2_data_processed |> 
  mutate(seropositive = case_when(interpretation == "Positive" ~ TRUE,
                                  TRUE ~ FALSE),
         household_id = str_remove(id, "-\\d+$"),
         # Collapse non-christian religions
         religion = fct_collapse(religion,
                                 "Christian" = "Christian",
                                 "Non-christian" = c("Muslim", "Traditionalist", "Other")),
         # Collapse educational levels
         education = fct_collapse(education,
                                  "None/No formal schooling" = "None/No formal schooling",
                                  "Primary" = "Primary education (up until age 12)",
                                  "Junior secondary" = "Junior secondary education (up until age 16)",
                                  "Senior secondary or greater" = c("Senior secondary education (up until age 19)", "University level education", "Master's, Doctoral level degree"),
                                  "Other" = "Other"))

hh_data_models <- tbl_1_data |>
         mutate(positive_hh = case_when(n_positive == 0 ~ FALSE,
                                        n_positive >= 1 ~ TRUE))

#source(here("main", "09_univariable_risk_factor_analysis.R"))

demographic_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "demographic_models_results.rds"))
environmental_hh_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "environmental_hh_results.rds"))
environmental_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "environmental_uni_results.rds"))
behavioural_hh_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "behavioural_hh_results.rds"))
behavioural_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "behavioural_uni_results.rds"))
occupational_models_results <- read_rds(here("manuscript", "cross_sectional_study_outputs", "models", "occupational_uni_results.rds"))

all_risk_factors_summary <- bind_rows(
  demographic_models_results$summary_table |> mutate(domain = "Demographic"),
  environmental_hh_models_results$summary_table |> mutate(domain = "Environmental (Household)"),
  environmental_models_results$summary_table |> mutate(domain = "Environmental (Individual)"),
  behavioural_hh_models_results$summary_table |> mutate(domain = "Behavioural (Household)"),
  behavioural_models_results$summary_table |> mutate(domain = "Behavioural (Individual)"),
  occupational_models_results$summary_table |> mutate(domain = "Occupational")
) |>
  mutate(domain = case_when(predictor == "community_pob" ~ "Demographic",
                            TRUE ~ domain))

```

Univariable models adjusting for village-level clustering found no strong evidence of association between seropositivity and most pre-specified risk factors (Figure 2). Odds ratios were generally close to 1.0 with wide intervals. Given these results and the study’s power limitations from overall low seropositivity, multivariable modelling was deemed unnecessary, we therefore only report univariable results below.

```{r}
#| label: produce-risk-factors
#| echo: false
#| warning: false

main_table_predictors <- c(
  # From Table 1
  "n_people", "n_buildings", "n_single_room", "proximity_bush", 
  "proximity_farm", "toilet_use", "rodents_enter",
  # From Table 2
  "age", "sex", "education", "community_pob", "farming", "ag_worker", "trader",
  "current_rodent_consumption", "past_rodent_consumption", "excreta_cleaning"
)

clean_labels <- c(
  "age" = "Age (per year increase)",
  "sexMale" = "Sex: Male vs Female",
  "educationPrimary" = "Education: Primary vs None",
  "educationSecondary" = "Education: Secondary vs None",
  "educationOther" = "Education: Post-secondary vs None",
  "community_pobYes" = "Born in current village: Yes vs No",
  "farming" = "Occupation: Farmer",
  "ag_worker" = "Occupation: Agricultural worker",
  "trader" = "Occupation: Trader",
  "current_rodent_consumptionYes" = "Current rodent consumption: Yes vs No",
  "past_rodent_consumptionYes" = "Past rodent consumption: Yes vs No",
  "excreta_cleaningYes" = "Cleaned rodent excreta: Yes vs No",
  "n_people" = "Household size (per person)",
  "n_buildings" = "Number of buildings (per building)",
  "n_single_room" = "Number of single-room buildings",
  "proximity_bushYes" = "Proximity to bush: Yes vs No",
  "proximity_farmYes" = "Proximity to farm: Yes vs No",
  "toilet_usePitlatrine" = "Toilet: Pit latrine vs Plumbed",
  "toilet_useFielddefecation" = "Toilet: Field defecation vs Plumbed",
  "rodents_enterYes" = "Rodents enter home: Yes vs No"
  )

main_plot_data <- all_risk_factors_summary |>
  filter(predictor %in% main_table_predictors) |>
  # Remove reference levels and unstable estimates
  filter(median_OR != 1, (log(upper) - log(lower)) < 20) |>
  mutate(variable_label = recode(term, !!!clean_labels),
         domain = fct(domain, levels = c("Demographic", "Environmental (Household)", "Environmental (Individual)", "Behavioural (Individual)", "Occupational")))

main_plot_figure <- main_plot_data |>
  ggplot(aes(x = median_OR, y = fct_reorder(variable_label, median_OR))) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
    geom_point(aes(color = domain), size = 2.5) +
    geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.15) +
    scale_x_log10(breaks = c(0.2, 0.5, 1, 2, 5)) +
    facet_grid(domain ~ ., scales = "free_y", space = "free_y") +
    labs(
      x = "Odds Ratio (log scale)",
      y = NULL
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      strip.text.y = element_text(angle = 0)
    )

save_plot(filename = here("output", "figure_2.pdf"), plot = main_plot_figure, base_width = 7, base_height = 6)

```

```{r}
#| label: plot-risk-factors
#| fig-cap: "Forest plot of univariable odds ratios for Lassa virus seropositivity. Points are posterior medians, and horizontal bars are 95% credible intervals, adjusted for clustering by village."
#| echo: false

knitr::include_graphics(here("output", "figure_2.pdf"))
```

### Demographic correlates of LASV seropositivity

No meaningful association was found for age (OR: 1.01 [0.99–1.02]) or sex (OR: 1.06 [0.64–1.76]). Post-secondary education showed a protective trend (OR: 0.39 [0.17–0.90]), though likely confounded by age and socioeconomic status.

### Environmental and Behavioral correlates of LASV seropositivty

No clear associations were found for household-level variables such as household size (OR: 0.97, [0.89–1.06]), number of buildings (OR: 0.80, [0.62–1.00]), rodent entry into the home (OR: 0.72, [0.27–2.30]), or cleaning of rodent excreta (OR: 0.42 [0.06–9.64]). Odds ratios for rodent consumption (OR: 0.93 [0.49–1.83]) and past rodent consumption only (OR: 0.59 [0.23–1.46]) were also uncertain.

## Spatial heterogeneity in LASV exposure

```{r}
#| label: spatial-analysis
#| echo: false
#| warning: false
#| execute: false

final_spatial_plot <- read_rds(file = here("output", "spatial_analysis", "spatial_plot.rds"))

save_plot(here("output", "figure_3.pdf"), plot = final_spatial_plot, base_width = 12, base_height = 12)
```

```{r}
#| label: spatial-fig
#| fig-cap: "Spatial distribution of household LASV seropositivity and Local Gi* clustering. Each panel represents a single village, showing jittered household locations. Shapes indicate whether at least one LASV-seropositive individual was recorded in the household. The color classification reflects the local clustering of seropositivity, where the Gi* test includes the household's own value in the local average; a high-value central household may thus be classified as 'Not significant' if surrounding seronegative households dilute its local average below the statistical threshold. Scale bars are included to indicate spatial scale; coordinates are not shown to preserve privacy."
#| echo: false
#| warning: false

knitr::include_graphics(here("output", "figure_3.pdf"))
```

We assessed the spatial clustering of LASV seropositivity at the household level across nine villages. Global spatial autocorrelation, assessed using Moran’s I, found no evidence of significant village-wide clustering in any site (all *p* > 0.05). Localised clustering was then evaluated using the Getis-Ord Gi* statistic, which identified a total of 27 hotspot and 6 coldspot households across all villages (Figure 3). However, these statistical clusters did not align well with observed seropositivity; of the 59 seropositive households in the study, only 3.6% were located within a designated hotspot.

Logistic regression confirmed no significant difference in the odds of seropositivity based on cluster status (*p* = 0.39 $OR_{clustered~vs.~non-clustered}$ = 0.53 [0.12-2.3]). These results strongly indicate that, in our study sites, seropositive households were not concentrated in discernible spatial clusters.

# Discussion

This study found that LASV exposure in rural Nigeria was substantially lower than anticipated based on national risk models [@reddingEnvironmentalmechanisticModellingImpact2016]. While the overall model-based IgG seroprevalence was 3.2% [2.5–4.0%], this aggregate figure masks marked heterogeneity at the local scale. Prevalence varied significantly between states (1.7% to 5.1%) and, more critically, between villages (0.8% to 6.5%). These findings challenge broad-scale risk mapping, pointing to a transmission landscape that is generally low but punctuated by complex, hyper-local variation supported by three observations: divergent age patterns, absence of risk factor correlation, and lack of spatial clustering.

Our seroprevalence estimates are comparable to those from some contemporary studies in West Africa, though such comparisons are conditioned by differences in study populations and laboratory methods [@Grant2023-em; @lerchProjectingVaccineDemand2022]. The high specificity of the commercial ELISA kit (LASV vs. non-LASV arenaviruses) used in this study may yield more conservative estimates than alternative serological assays [@Soubrier2022-gr]. More revealing was the relationship with age. Two villages exhibited classic increasing seroprevalence with age, indicative of cumulative, endemic transmission. Conversely, two others showed peaks in younger groups, suggesting dynamic local transmission [@haySerodynamicsPrimerSynthetic2024]. 

Peaks in younger age groups could arise from recent increases in transmission or immunological factors like seroreversion. With antibody waning estimated at ~3% annually [@mooreEstimationLassaFever2025], the decline in older cohorts likely reflects the loss of detectable markers from historical infections. This implies that cross-sectional seroprevalence may serve only as a moving window of recent exposure (e.g., the last 15–20 years) rather than a cumulative record of lifetime risk. Consequently, the distinct peak in younger age groups observed in villages like Okimbongha [CR] likely signals intense, recent transmission events concentrated in younger cohorts, while the low prevalence in older adults reflects both lower recent exposure and the decay of historical antibodies. From a trial planning perspective, these communities with steep age-seroprevalence gradients in younger cohorts likely represent the most viable candidates for efficacy studies, as they indicate active, ongoing viral circulation (high force of infection) rather than historical exposure.

Consistent with this theme of local heterogeneity, our univariable analyses revealed no strong or consistent demographic, environmental, or occupational correlates of seropositivity. The lack of clear associations with commonly cited risks, such as farming, contrasts with some previous findings [@ogundeleLassaFeverWest2025]. This may be a function of the study's limited statistical power, a consequence of the low overall seroprevalence, to detect modest individual-level effects. Furthermore, the near-universal reporting of certain exposures, such as the presence of rodents in homes (94%), creates little variability. This complicates the quantification of their true impact, particularly as risk is ultimately driven by the prevalence of LASV within local rodent populations, which itself can be highly variable, again highlighting the heterogeneous nature of risk [@marienHouseholdsHotspotsLassa2020; @banguraLassaVirusCirculation2021]. 

The lack of association with rodent consumption, despite its high prevalence in BN and CR (up to 91%), challenges the assumption that hunting or consumption is the primary driver of spillover. It suggests that transmission may occur more frequently through indirect routes, such as contamination of food stores or household surfaces by rodent urine, which affects the entire household regardless of who consumes the meat

Collectively, these findings suggest a departure from a uniform transmission model, indicating instead a process that is intensely local and patchy. It is plausible that LASV exposure risk is governed by a complex interplay of factors at the micro-scale of the household or individual, rather than being broadly structured by village geography or demographics. Such factors may include unmeasured variables like household-specific rodent density and viral shedding status, subtle differences in food storage and sanitation practices, or individual activities in high-risk locations not fully captured by our questionnaire.

## Strengths and Limitations

The strengths of this study include its systematic, multi-criteria site selection; a large sample size across diverse communities; and a robust analytical framework using Bayesian mixed models to account for data hierarchies and uncertainty. The inclusion of formal spatial statistics adds rigor to our conclusion that transmission is not strongly clustered at the household level. Nevertheless, the study has important limitations. It was powered to detect village-level differences based on higher seroprevalence estimates from previous studies; however, the observed low seroprevalence limited our power for individual-level risk factor analysis. 

The cross-sectional design precludes assessment of causality and the distinction between recent and past infections. Consequently, while baseline seroprevalence is a superior screening tool compared to regional risk maps, it remains a proxy for true incidence. Ideally, serosurveys should be used to stratify villages for subsequent prospective incidence studies. This challenge is compounded by the potential for antibody waning and sero-reversion, the rates of which are critical for accurate Force of Infection estimation but remain poorly understood for LASV. Furthermore, the measurement of behaviour and environment at a single cross-sectional point may not accurately reflect cumulative lifetime exposure, as housing and habits vary over time. Finally, reliance on self-reported behaviours is subject to recall bias, and the study lacked objective ecological measures, such as rodent abundance data.

## Public Health Implications and Future Directions

The absence of spatial clustering and a clear risk profile suggests that geographically-targeted interventions may be inefficient. Furthermore, these findings highlight a critical vulnerability in the planning of Phase III vaccine efficacy trials. Our data demonstrate that being located within a high-risk state or Local Government Area does not guarantee high local exposure. Consequently, trial sites selected based on regional notification rates may inadvertently target villages with insufficient transmission intensity to demonstrate vaccine efficacy. To mitigate this risk, clinical trial site selection must adopt a One Health approach, moving beyond passive surveillance data to incorporate granular, active human serosurveys alongside prospective, longitudinal monitoring of rodent reservoirs. This integrated strategy ensures that trials are enrolled in locations where the virus is demonstrably circulating in the animal host, thereby maximising the likelihood of capturing spillover events.

Public health efforts should prioritize broadly-deployed, household-level interventions (e.g., rodent-proofing) and maintain high suspicion in surveillance across all villages within endemic regions, not just sentinel sites. Surveillance must also include areas of previously low-predicted risk (e.g., cold spots) given the observed micro-heterogeneity. Future research must employ a One Health approach, leveraging data from our ongoing longitudinal rodent sampling [@friantScalingLassaVirus2025]. Ultimately, disentangling these patterns requires longitudinal cohort studies linking exposures with seroconversion.

## Conclusion

LASV exposure in rural Nigeria is highly variable and likely driven by complex, hyper-local factors. These findings caution against reliance on broad-scale risk mapping for public health planning. Effective control requires shifting toward adaptable, household-focused interventions and, critically, the adoption of granular, One Health surveillance strategies to accurately identify high-transmission sites for clinical research and vaccine deployment.

# Acknowledgments

We sincerely thank the residents and leadership of the nine study communities in Benue, Cross River, and Ebonyi states for their warm welcome, hospitality, and willingness to participate in this research. We are deeply grateful to the community liaisons and village guides whose local knowledge and dedication were vital for facilitating community entry and navigation. We also acknowledge the State Ministries of Health and the local public health teams, including the Disease Surveillance and Notification Officers (DSNOs), for their essential logistical support and ongoing collaboration throughout the fieldwork. Funding was provided by the joint NSF‐NIH‐NIFA Ecology and Evolution of Infectious Disease Award #2208034 in partnership with Research and Innovation (UKRI) Biotechnology and Biological Sciences Research Council (BBSRC) Award BB/X005364/1.

# References
